---
title: "git@github.com:amesclir/Ecoflor2017_WorkShop_BasicPhylogenetics"
author: "Marcial"
date: "01/16/2017"
output: html_document
---

0. Visit this website: http://ape-package.ird.fr/APER.html
You can find materials in this website.


1. Let's download the GenBank sequences

```{r}

library(ape)


x <- paste("AJ5345", 26:49, sep = "")
x <- c("Z73494", x)
x
sylvia.seq <- read.GenBank(x)

#Problems with https in linux? Use this!
library(RCurl)
source("read.GenBank2.R")
sylvia.seq <- read.GenBank2(x)
sylvia.seq

#Exploring the data
names(sylvia.seq)
attributes(sylvia.seq)
attr(sylvia.seq, "names")
attr(sylvia.seq, "species")
attr(sylvia.seq, "description")
attr(sylvia.seq, "class")

save(list = ls(all.names = TRUE), file = "sylvia.RData")
```

2. Alignment of sequences
```{r}
sylvia.clus <- clustal(sylvia.seq)
#clustal is an older method
sylvia.clus

library(phyloch)
sylvia.muscle <- muscle(sylvia.seq)
#muscle is a more updated method
sylvia.muscle

identical(sylvia.clus[x, ], sylvia.muscle[x, ])
#this compare both alignments

save(list = ls(all.names = TRUE), file = "sylvia.RData")
```

3. Edit sequence data, load ecological data and save data
```{r}
taxa.sylvia <- attr(sylvia.seq, "species")
taxa.sylvia
# We have created an object with the scientific names of the species of the downloaded sequences
names(taxa.sylvia) <- names(sylvia.seq)
#We use the GenBank accession number to match list of scientific names and their sequences
taxa.sylvia[1] <- "Sylvia_atricapilla"
taxa.sylvia[24] <- "Sylvia_abyssinica"
taxa.sylvia[20] <- "Sylvia_melanocephala"
taxa.sylvia[15] <- "Sylvia_conspicillata"
# we have to update some scientific names

sylvia.eco <- read.table("sylvia_data.txt")
sylvia.eco
str(sylvia.eco)
rownames(sylvia.eco)

save(list = ls(all.names = TRUE), file = "sylvia.RData")

```



4. DNA models of evolution
Visit this website http://svitsrv25.epfl.ch/R-doc/library/ape/html/dist.dna.html

```{r}

load (file = "sylvia.RData")
sylvia.seq.ali <- sylvia.muscle
syl.K80 <- dist.dna(sylvia.seq.ali, pairwise.deletion = TRUE)
#K80 transitions (A <-> G, i.e. from purine to purine, or C <-> T, i.e. from pyrimidine to pyrimidine) and transversion and equal base frequencies.
syl.F84 <- dist.dna(sylvia.seq.ali, model = "F84", p = TRUE)
#K84 transitions (A <-> G, i.e. from purine to purine, or C <-> T, i.e. from pyrimidine to pyrimidine) and transversion and unequal base frequencies.
syl.TN93 <- dist.dna(sylvia.seq.ali, model = "TN93", p = TRUE)
#TN93 2 transition rates (A <-> G, i.e. from purine to purine, or C <-> T, i.e. from pyrimidine to pyrimidine) and 1 transversion rate and unequal base frequencies.
syl.GG95 <- dist.dna(sylvia.seq.ali, model = "GG95", p = TRUE)
#Galtier and Gouy (1995) introduced a model where the G+C content may change through time. Different rates are assumed for transitons and transversions.
round(cor(cbind(syl.K80, syl.F84, syl.TN93, syl.GG95)), 3)

syl.JC69 <- dist.dna(sylvia.seq.ali, model = "JC69", p = TRUE)
# One rate and equal base frequencies
syl.raw <- dist.dna(sylvia.seq.ali, model = "raw", p = TRUE)

layout(matrix(1:5, 1))
plot(syl.JC69, syl.raw)
abline(b = 1, a = 0) # draw x = y line
plot(syl.K80, syl.JC69)
abline(b = 1, a = 0)
plot(syl.F84, syl.K80)
abline(b = 1, a = 0) # draw x = y line
plot(syl.TN93, syl.F84)
abline(b = 1, a = 0)
plot(syl.F84, syl.GG95)
abline(b = 1, a = 0)

#Let's find which is the best DNA evolution model for our matrix
library(phangorn)
sylvia.seq.ali.phyDat <- as.phyDat(sylvia.seq.ali)
sylvia.seq.ali.phyDat
modelTest.results <- modelTest(sylvia.seq.ali.phyDat, tree=NULL, model = c("JC", "F81", "K80", "HKY", "SYM", "GTR"), G = TRUE, I = TRUE, FREQ = FALSE, k = 4, control = pml.control(epsilon = 1e-08,
maxit = 10, trace = 1), multicore = FALSE, mc.cores = NULL)
#JC one parameter
#F81 one parameter for tranv. and trans. and equal base frequences
#K80 two parameters, tranv. vs. trans.
#HKY two parameters, tranv. vs. trans. and unequal base frequences
#SYM six parameeters, and equal base frequences
#GTR six parameters, equal base frequences
modelTest.results

save(list = ls(all.names = TRUE), file = "sylvia.RData")

```


5. Making NJ trees based on distance matrices
```{r}

layout(matrix(1:1, 1))
nj.sylvia.K80 <- nj(syl.K80)
plot(nj.sylvia.K80)

nj.sylvia.K80$tip.label<-taxa.sylvia[nj.sylvia.K80$tip.label==names(taxa.sylvia)]
plot(nj.sylvia.K80)

nj.sylvia.K80 <- root(nj.sylvia.K80, "Chamaea_fasciata")
plot(nj.sylvia.K80)

nj.sylvia.GG95 <- nj(syl.GG95)
nj.sylvia.GG95$tip.label<-taxa.sylvia[nj.sylvia.GG95$tip.label==names(taxa.sylvia)]
nj.sylvia.GG95 <- root(nj.sylvia.GG95, "Chamaea_fasciata")
plot(nj.sylvia.GG95)


dist.topo(nj.sylvia.K80, nj.sylvia.GG95)

save(list = ls(all.names = TRUE), file = "sylvia.RData")
```

6. Bootstrap analyses in a NJ tree

```{r}

grep("Chamaea", taxa.sylvia, value = TRUE)
f <- function(xx) root(nj(dist.dna(xx, p=TRUE)), "AJ534526")
tr <- f(sylvia.muscle)
tr
plot(tr)
## same than: tr <- root(nj.sylvia.K80, "AJ534526")

nj.boot.sylvia <- boot.phylo(tr, sylvia.muscle, f, 200,
                             rooted = TRUE)
#deffault is K80
nj.boot.sylvia


nj.est <- tr
nj.est$tip.label <- taxa.sylvia[tr$tip.label]
plot(nj.est, no.margin = TRUE)
nodelabels(round(nj.boot.sylvia / 200, 2), bg = "white")
add.scale.bar(length = 0.01)
write.tree(nj.est, "sylvia_nj_k80.tre")

save(list = ls(all.names = TRUE), file = "sylvia.RData")

```


7. Maximum likelihood analyses

```{r}
#ML with phyml
write.dna(sylvia.muscle, "sylvia.phy")
#This saves the file in phy format
phyml.sylvia <- phymltest("sylvia.phy", execname = "~/local/bin/PhyML-3.1/PhyML-3.1_linux64")
summary(phyml.sylvia)
plot(phyml.sylvia, col = "black")
TR <- read.tree("sylvia.phy_phyml_tree.txt")
TR
names(TR)
mltr.sylvia <- TR[[28]]
mltr.sylvia$tip.label <- taxa.sylvia[mltr.sylvia$tip.label]
mltr.sylvia <- root(mltr.sylvia, "Chamaea_fasciata")
mltr.sylvia <- drop.tip(mltr.sylvia, "Chamaea_fasciata")
plot(mltr.sylvia, no.margin = TRUE)
add.scale.bar(length = 0.01)

write.dna(sylvia.muscle, "sylvia2.phy")
phyml.sylvia.GTRIG <- phymltest("sylvia2.phy", exclude = c("JC69", "K80", "F81", "F84", "HKY85", "TN93", "GTR", "JC69+I", "K80+I", "F81+I", "F84+I", "HKY85+I", "TN93+I", "GTR+I", "JC69+G", "K80+G", "F81+G", "F84+G", "HKY85+G", "TN93+G", "GTR+G", "JC69+I+G", "K80+I+G", "F81+I+G", "F84+I+G", "HKY85+I+G", "TN93+I+G"), execname = "~/local/bin/PhyML-3.1/PhyML-3.1_linux64")
TRGTRIG <- read.tree("sylvia2.phy_phyml_tree.txt")

#ML with RAXML
library(phyloch)
raxml.sylvia.GTRCAT <- raxml(sylvia.seq.ali, path = "~/local/bin/RAxML-7.0.4/")
#Must be this version of raxml. If you use other newer version, it won't work
setwd("/home/amesclir/Documents/Ubuntufiles/GITHUB/Ecoflor2017_WorkShop_BasicPhylogenetics")

raxml.sylvia.GTRCAT$tip.label <- taxa.sylvia[raxml.sylvia.GTRCAT$tip.label]
raxml.sylvia.GTRCAT <- root(raxml.sylvia.GTRCAT, "Chamaea_fasciata")
raxml.sylvia.GTRCAT <- drop.tip(raxml.sylvia.GTRCAT, "Chamaea_fasciata")
plot(raxml.sylvia.GTRCAT, no.margin = TRUE)
add.scale.bar(length = 0.01)
write.tree(raxml.sylvia.GTRCAT, file = "raxml.sylvia.GTRCAT.tree")


#ML with function implemented in phangorn
library(phangorn)
  dmJC <- dist.ml(sylvia.seq.ali.phyDat)
  tree <- NJ(dmJC)
  tree <- root(tree, "AJ534526")
# Jukes-Cantor (starting tree from NJ)  
  fitJC <- pml(tree, sylvia.seq.ali.phyDat)
fitJC
fitJC$tree
# optimize edge length parameter     
  fitJC <- optim.pml(fitJC)
  fitJC
fitJC$tree
## Not run:     
# search for a better tree using NNI rearrangements     
  fitJC <- optim.pml(fitJC, optNni=TRUE)
  fitJC
fitJC$tree
# JC + Gamma + I - model
fitJC_GI <- update(fitJC, k=4, inv=.2)
fitJC_GI
# optimize shape parameter + proportion of invariant sites     
fitJC_GI <- optim.pml(fitJC_GI, optGamma=TRUE, optInv=TRUE)
fitJC_GI
# GTR + Gamma + I - model
  fitGTRGI <- optim.pml(fitJC_GI, rearrangement = "stochastic", 
      optGamma=TRUE, optInv=TRUE, model="GTR") 

fitGTRGI_def <- optim.pml(fitJC_GI, rearrangement = "NNI", 
      optGamma=TRUE, optInv=TRUE, model="GTR") 

fitGTRGI_def2 <- fitGTRGI_def
fitGTRGI_def2$tree <- root(fitGTRGI_def2$tree, "AJ534526")
fitGTRGI_def2$tree$tip.label <- taxa.sylvia[fitGTRGI_def2$tree$tip.label]
  plot(fitGTRGI_def2$tree, cex = 0.8)
write.tree(fitGTRGI_def2$tree, file = "optim.plm.sylvia.GTRIG.tree")

save(list = ls(all.names = TRUE), file = "sylvia.RData")

```

8. Bootstrap analysis and maximum likelihood

```{r}
#Bootstrap in phyml not available for R

#Bootstrap in raxml
library(phyloch)
raxml.sylvia.GTRCAT.boot <- raxml2(sylvia.seq.ali, runs = c(2, 5), outgroup = "AJ534526", path = "~/local/bin/RAxML-7.0.4/")
setwd("/home/amesclir/Documents/Ubuntufiles/GITHUB/Ecoflor2017_WorkShop_BasicPhylogenetics")
#There is a bug in this old version of raxml and the bootstrap analysis does not work, a least for this data set

# Let's run it with the terminal using this line
write.dna(sylvia.muscle, "~/local/bin/RAXML2/sylvia.phy")
#Use this command 
#raxmlHPC -f a -p 12345 -s sylvia.phy -x 12345 -# 100 -m GTRCAT -n sylvia.raxml
raxml.sylvia.GTRCAT.boot <- read.tree("~/local/bin/RAXML2/RAxML_bipartitions.sylvia.raxml")

raxml.sylvia.GTRCAT.boot$tip.label <- taxa.sylvia[raxml.sylvia.GTRCAT.boot$tip.label]
raxml.sylvia.GTRCAT.boot <- root(raxml.sylvia.GTRCAT.boot, "Chamaea_fasciata")
raxml.sylvia.GTRCAT.boot <- drop.tip(raxml.sylvia.GTRCAT.boot, "Chamaea_fasciata")
plot(raxml.sylvia.GTRCAT.boot, no.margin = TRUE, show.node.label=T)
add.scale.bar(length = 0.01)
setwd("/home/amesclir/Documents/Ubuntufiles/GITHUB/Ecoflor2017_WorkShop_BasicPhylogenetics")



#Bootstrap in function implemented in phangorn
set.seed(123)

bs <- bootstrap.pml(fitGTRGI_def, bs=100, optNni=TRUE)
fitGTRGI_def$tree <- root(fitGTRGI_def$tree, "AJ534526")
treeBS <- plotBS(fitGTRGI_def$tree,bs,type="phylogram")
ml.boot.sylvia <- prop.clades(fitGTRGI_def$tree, bs, part = NULL, rooted = FALSE)
mytree <- fitGTRGI_def$tree
mytree$tip.label <- taxa.sylvia[mytree$tip.label]
mytree <- drop.tip(mytree, "Chamaea_fasciata")
plot(mytree, no.margin = TRUE)
nodelabels(round(ml.boot.sylvia / 100, 2), bg = "white")
add.scale.bar(length = 0.01)
write.tree(nj.est, "sylvia_bootstrap.pml_gtrig.tre")

```

9. Phylogenetic Bayesian analysis

```{r}
library(phyloch)

setwd("/home/amesclir/local/bin/mrbayes-3.1.2")
mrbayesresults <- mrbayes(sylvia.seq.ali, file = "sylvia_bayes.nex", nst = 6, rates = "invgamma", ngammacat = 4,     nruns = 2, ngen = 1e+05, printfreq = 100, samplefreq = 100,     nchains = 4, savebrlens = "yes", temp = 0.2, burnin = 10, contype = "allcompat", run = TRUE)

sylvia.bayes.tree <- read.nexus("sylvia_bayes.nex.con")

setwd("/home/amesclir/Documents/Ubuntufiles/GITHUB/Ecoflor2017_WorkShop_BasicPhylogenetics")
write.tree(sylvia.bayes.tree, "sylvia_bayes.nex.con" )
sylvia.bayes.tree <- read.tree("sylvia_bayes.nex.con")

sylvia.bayes.tree[[1]] <- root(sylvia.bayes.tree[[1]], "AJ534526")
sylvia.bayes.tree[[1]] <- drop.tip(sylvia.bayes.tree[[1]], "AJ534526")
sylvia.bayes.tree[[1]]$tip.label <- taxa.sylvia[sylvia.bayes.tree[[1]]$tip.label]
plot(sylvia.bayes.tree[[1]], no.margin = TRUE, show.node.label=T)


save(list = ls(all.names = TRUE), file = "sylvia.RData")
```


10. chronogram with the penalized likelihood method and bayesian approach

```{r}
#dating using penalized likelihood method
tr.ml <- drop.tip(mltr.sylvia, "Chamaea_fasciata")
res <- vector("list", 9)
for (L in -4:4)
    res[[L + 5]] <- chronopl(tr.ml, 10^L, 12, 16, CV = TRUE)
Lambda <- 10^(-4:4)
CV <- sapply(res, function(x) sum(attr(x, "D2")))

layout(matrix(1:1, 1))
par(mai=c(1.02,0.82,0.82,0.42))    

plot(Lambda, CV / 1e5, log = "x")

sylvia.chrono <- res[[2]]
rts <- attr(sylvia.chrono, "rates")
summary(rts)

par(mar = c(2, 0, 0, 0))
plot(sylvia.chrono, edge.width = 100*rts, label.offset = .15)
axisPhylo()

write.tree(sylvia.chrono, "sylvia.chrono.tree")



######################
#dating using lagopus#
######################
#This is not actually working.
#Usa beast better http://beast2.org/

multidivtime.sylvia <- multidivtime(mdt.in.sylvia, file = NULL, start = "baseml", part = 1,
  runs = 1, path = path, transfer.files = TRUE, 
	LogLCheck = 100, plot = TRUE)

setwd("/home/amesclir/Documents/Ubuntufiles/GITHUB/Ecoflor2017_WorkShop_BasicPhylogenetics")



## Create object of class 'mdt.in'.
library(LAGOPUS)
sylvia.bayes.tree[[2]] <- root(sylvia.bayes.tree[[2]], "AJ534526")
plot(sylvia.bayes.tree[[2]])

plot(sylvia.bayes.tree[[2]])
node.A <- c("L", "AJ534548", "AJ534529", 12)
node.B <- c("U", "AJ534548", "AJ534529", 16)
my.age.constraints <- data.frame(rbind(node.A, node.B))

mdt.in.sylvia <- mdt.in(sylvia.seq.ali, sylvia.bayes.tree[[2]],my.age.constraints)

## Load defaults for baseml and multidivtime control files.
data(baseml.ctr)
data(multicntrl.dat)

## Have a look at '?baseml.ctr' and 'multicntrl.dat' for how
## to change the defaults.

## Not run:
## Specify a path to the executables.
path <- "/home/amesclir/local/bin/multidistribute/"

multidivtime(mdt.in.sylvia, file = NULL, start = "baseml", part = 1,
  runs = 1, path = path, transfer.files = TRUE, 
	LogLCheck = 100, plot = TRUE)
## Run another MCMC on the same data set, but with
multidivtime(mdt.in.sylvia, file = NULL, start = "multidivtime", part = 1,
  runs = 1, path = path, transfer.files = TRUE, 
  LogLCheck = 100, plot = TRUE)

##Not working. Maybe the version has changed.

```



11. Ancestral character reconstruction. Discrete characters.

```{r}

plot(sylvia.chrono)
DF <- sylvia.eco[sylvia.chrono$tip.label, ]
table(DF$geo.range, DF$mig.behav)

syl.er <- ace(DF$geo.range, sylvia.chrono, type = "d")
syl.er
syl.sym <- ace(DF$geo.range, sylvia.chrono, type="d", model="SYM")
syl.sym
anova(syl.er, syl.sym)
sapply(list(syl.er, syl.sym), AIC)

mod <- matrix(0, 3, 3)
mod[2, 1] <- mod[1, 2] <- 1
mod[2, 3] <- mod[3, 2] <- 2
mod
syl.mod <- ace(DF$geo.range, sylvia.chrono, type="d", model=mod)
syl.mod
sapply(list(syl.er, syl.sym, syl.mod), AIC)

Q <- syl.mod$index.matrix
Q
diag(Q) <- 0
Q
Q[1, 2] <- Q[2, 1] <- syl.mod$rates[1]
Q[2, 3] <- Q[3, 2] <- syl.mod$rates[2]
Q

Q <- syl.mod$index.matrix
diag(Q) <- 0
Q
Q[] <- c(0, syl.mod$rates)[Q + 1]
Q
diag(Q) <- -rowSums(Q)
Q


P <- matexpo(1 * Q)
#This function computes the exponential of a square matrix using a spectral decomposition. 
rownames(P) <- c("temp", "temptrop", "trop")
colnames(P) <- rownames(P)
P

co <- rep("grey", 24)
co[DF$geo.range == "temp"] <- "black"
co[DF$geo.range == "trop"] <- "white"
plot(sylvia.chrono, type = "c", use.edge.length = FALSE, no.margin = TRUE, label.offset = 1)
tiplabels(pch = 22, bg = co, cex = 2, adj = 1)
nodelabels(thermo = syl.mod$lik.anc, cex = 0.8,
           piecol = c("black", "grey", "white"))


```

12. Diversification analyses
```{r}

#diversification analyses with ape
sylvia.chrono <- read.tree("sylvia.chrono.tree")
yule(sylvia.chrono)
birthdeath(sylvia.chrono)
1 - pchisq(2*(-7.858507 - -7.858507), 1)


#Diversification analyses with laser
library(laser)
library(phytools)

mytree <- read.tree("sylvia.chrono.tree")

layout(matrix(1:1, 1))
par(mai=c(1.02,0.82,0.82,0.42))    

mytreeltt <- ltt(mytree, plot=T, drop.extinct=FALSE, log.lineages=TRUE, gamma=TRUE)
mytreeltt

myBtimes <- getBtimes("sylvia.chrono.tree")

myfitdAICrc <- fitdAICrc(myBtimes, modelset = c("pureBirth", "bd", "DDL", "DDX", "yule2rate", "yule3rate"), ints = 100)
myfitdAICrc


```


13. Diversification and character states

```{r}

#test for different diversification rates for different character state using yule.cov
x <- sylvia.eco[sylvia.chrono$tip.label, "geo.range"]
ANC <- ace(x, sylvia.chrono, type = "d", model = mod)
ANC$lik.anc[1:3, ]
anc <- apply(ANC$lik.anc, 1, which.max)
anc

X <- factor(c(x, anc))
yule.cov(sylvia.chrono, ~ X)

1 / (1 + exp(-(0.3200766)))
1 / (1 + exp(-(0.3200766 -2.0038455)))
1 / (1 + exp(-(0.3200766 -1.8433086)))

#bootstrapping
source("yule.cov2.R")
yule.cov2(sylvia.chrono, ~ X)
fsamp <- function(x) sample(length(x), size = 1, prob = x)
nrep <- 1e3
Pvls <- numeric(nrep)
for (i in 1:nrep) {
    anc <- apply(ANC$lik.anc, 1, fsamp)
    X <- factor(c(x, anc))
    Pvls[i] <- yule.cov2(sylvia.chrono, ~ X)
}
hist(Pvls, freq = FALSE, main = "")
lines(density(Pvls))


#test for different diversification rates for different character states using GEOSSE in the library(diversitree)
library(diversitree)
phy <- sylvia.chrono
tip.state <- sylvia.eco[sylvia.chrono$tip.label, "geo.range"]
tip.state <- as.character(tip.state)
tip.state[tip.state=="temptrop"] <- 2
tip.state[tip.state=="temp"] <- 1
tip.state[tip.state=="trop"] <- 0
names(tip.state) <- sylvia.chrono$tip.label
tip.state

statecols <- c("1"="blue", "2"="purple", "0"="red")
plot(phy, tip.color=statecols[tip.state], cex=0.5)

p <- starting.point.geosse(phy)
p

tip.state <- as.numeric(tip.state)
names(tip.state) <- sylvia.chrono$tip.label

lik1 <- make.geosse(phy, tip.state)
lik2 <- constrain(lik1, sAB ~ 0)
lik3 <- constrain(lik1, sA ~ sB, xA ~ xB)

ml1 <- find.mle(lik1, p)
p <- coef(ml1)
ml2 <- find.mle(lik2, p[argnames(lik2)])
ml3 <- find.mle(lik3, p[argnames(lik3)])

round(rbind(full = coef(ml1),
no.sAB = coef(ml2, TRUE),
eq.div = coef(ml3, TRUE)), 3)

anova(ml1, no.sAB = ml2, eq.div = ml3)


geosse.asr <- asr.marginal(lik2, coef(ml2))

plot(phy,use.edge.length=T, label.offset=1.3, no.margin=TRUE, cex=0.7)
co <- c("0"="red", "1"="blue", "2"="purple")
tiplabels(pch=22, bg=co[as.numeric(tip.state + 1)], col=co[as.numeric(tip.state + 1)], cex=1, adj=1)
nodelabels(pie=t(geosse.asr), piecol = c("0"="red", "1"="blue", "2"="purple"), cex=0.7)
legend("bottomleft", c("Tropical","Temperate","Widespread"),pch=c(16,16),col=c("red", "blue","purple"),cex=0.7)

```









