---
title: "git@github.com:amesclir/Ecoflor2017_WorkShop_BasicPhylogenetics"
author: "Marcial"
date: "01/16/2017"
output: html_document
---

0. Visit this website: http://ape-package.ird.fr/APER.html
You can find materials in this website.


1. Let's download the GenBank sequences

```{r}

library(ape)

x <- paste("AJ5345", 26:49, sep = "")
x <- c("Z73494", x)
sylvia.seq <- read.GenBank(x)

#Problems with https in linux? Use this!
library(RCurl)
source("read.GenBank2.R")
sylvia.seq <- read.GenBank2(x)
sylvia.seq
names(sylvia.seq)
attributes(sylvia.seq)
attr(sylvia.seq, "names")
attr(sylvia.seq, "species")
attr(sylvia.seq, "description")
attr(sylvia.seq, "class")
```

2. Alignment of sequences
```{r}
sylvia.clus <- clustal(sylvia.seq)
#clustal is an older method
sylvia.clus

library(phyloch)
sylvia.muscle <- muscle(sylvia.seq)
#muscle is a more updated method
sylvia.muscle

identical(sylvia.clus[x, ], sylvia.muscle[x, ])
#this compare both alignments
```

3. Edit sequence data, load ecological data and save data
```{r}
taxa.sylvia <- attr(sylvia.seq, "species")
taxa.sylvia
# We have creaed an object with the scientific names of the species of the downloaded sequences
names(taxa.sylvia) <- names(sylvia.seq)
#We use the GenBank accession number to match list of scientific names and their sequences
taxa.sylvia[1] <- "Sylvia_atricapilla"
taxa.sylvia[24] <- "Sylvia_abyssinica"
taxa.sylvia[20] <- "Sylvia_melanocephala"
taxa.sylvia[15] <- "Sylvia_conspicillata"
# we have to update some scientific names

sylvia.eco <- read.table("sylvia_data.txt")
sylvia.eco
str(sylvia.eco)
rownames(sylvia.eco)
save(sylvia.muscle, taxa.sylvia, sylvia.eco,
     file = "sylvia.RData")
```



4. DNA models of evolution
Visit this website http://svitsrv25.epfl.ch/R-doc/library/ape/html/dist.dna.html

```{r}

load (file = "sylvia.RData")
sylvia.seq.ali <- sylvia.muscle
syl.K80 <- dist.dna(sylvia.seq.ali, pairwise.deletion = TRUE)
#K80 transitions (A <-> G, i.e. from purine to purine, or C <-> T, i.e. from pyrimidine to pyrimidine) and transversion and equal base frequencies.
syl.F84 <- dist.dna(sylvia.seq.ali, model = "F84", p = TRUE)
#K84 transitions (A <-> G, i.e. from purine to purine, or C <-> T, i.e. from pyrimidine to pyrimidine) and transversion and unequal base frequencies.
syl.TN93 <- dist.dna(sylvia.seq.ali, model = "TN93", p = TRUE)
#TN93 2 transition rates (A <-> G, i.e. from purine to purine, or C <-> T, i.e. from pyrimidine to pyrimidine) and 1 transversion rate and unequal base frequencies.
syl.GG95 <- dist.dna(sylvia.seq.ali, model = "GG95", p = TRUE)
#Galtier and Gouy (1995) introduced a model where the G+C content may change through time. Different rates are assumed for transitons and transversions.
round(cor(cbind(syl.K80, syl.F84, syl.TN93, syl.GG95)), 3)

syl.JC69 <- dist.dna(sylvia.seq.ali, model = "JC69", p = TRUE)
# One rate and equal base frequencies
syl.raw <- dist.dna(sylvia.seq.ali, model = "raw", p = TRUE)
layout(matrix(1:2, 1))
plot(syl.JC69, syl.raw)
abline(b = 1, a = 0) # draw x = y line
plot(syl.K80, syl.JC69)
abline(b = 1, a = 0)

library(phangorn)
sylvia.seq.ali.phyDat <- as.phyDat(sylvia.seq.ali)
sylvia.seq.ali.phyDat
modelTest.results <- modelTest(sylvia.seq.ali.phyDat, tree=NULL, model = c("JC", "F81", "K80", "HKY", "SYM", "GTR"), G = TRUE, I = TRUE, FREQ = FALSE, k = 4, control = pml.control(epsilon = 1e-08,
maxit = 10, trace = 1), multicore = FALSE, mc.cores = NULL)
#JC one parameter
#F81 one parameter for tranv. and trans. and equal base frequences
#K80 two parameters, tranv. vs. trans.
#HKY two parameters, tranv. vs. trans. and unequal base frequences
#SYM six parameeters, and equal base frequences
#GTR six parameters, equal base frequences
modelTest.results

```


5. Making NJ trees based on distance matrices
```{r}

layout(matrix(1:1, 1))
nj.sylvia.K80 <- nj(syl.K80)
plot(nj.sylvia.K80)

nj.sylvia.K80$tip.label<-taxa.sylvia[nj.sylvia.K80$tip.label==names(taxa.sylvia)]
plot(nj.sylvia.K80)

nj.sylvia.K80 <- root(nj.sylvia.K80, "Chamaea_fasciata")
plot(nj.sylvia.K80)

nj.sylvia.GG95 <- nj(syl.GG95)
nj.sylvia.GG95$tip.label<-taxa.sylvia[nj.sylvia.GG95$tip.label==names(taxa.sylvia)]
nj.sylvia.GG95 <- root(nj.sylvia.GG95, "Chamaea_fasciata")
plot(nj.sylvia.GG95)

dist.topo(nj.sylvia.K80, nj.sylvia.GG95)

```

6. Bootstrap analyses in a NJ tree

```{r}
grep("Chamaea", taxa.sylvia, value = TRUE)
f <- function(xx) root(nj(dist.dna(xx, p=TRUE)), "AJ534526")
tr <- f(sylvia.muscle)
tr
plot(tr)
## same than: tr <- root(nj.sylvia.K80, "AJ534526")

nj.boot.sylvia <- boot.phylo(tr, sylvia.muscle, f, 200,
                             rooted = TRUE)
nj.boot.sylvia


nj.est <- tr
nj.est$tip.label <- taxa.sylvia[tr$tip.label]
plot(nj.est, no.margin = TRUE)
nodelabels(round(nj.boot.sylvia / 200, 2), bg = "white")
add.scale.bar(length = 0.01)
write.tree(nj.est, "sylvia_nj_k80.tre")

```


7. Maximum likelihood analyses
#HASTA AQUI
```{r}
write.dna(sylvia.muscle, "sylvia.txt")
phyml.sylvia <- phymltest("sylvia.txt", execname = "~/local/bin/PhyML-3.1/PhyML-3.1_linux64")
summary(phyml.sylvia)
plot(phyml.sylvia, col = "black")
TR <- read.tree("sylvia.txt_phyml_tree.txt")
mltr.sylvia <- TR[[28]]
mltr.sylvia$tip.label <- taxa.sylvia[mltr.sylvia$tip.label]
mltr.sylvia <- root(mltr.sylvia, "Chamaea_fasciata")
plot(mltr.sylvia, no.margin = TRUE)
add.scale.bar(length = 0.01)

```

7. chronogram with the penalized likelihood method

```{r}

tr.ml <- drop.tip(mltr.sylvia, "Chamaea_fasciata")
res <- vector("list", 9)
for (L in -4:4)
    res[[L + 5]] <- chronopl(tr.ml, 10^L, 12, 16, CV = TRUE)
Lambda <- 10^(-4:4)
CV <- sapply(res, function(x) sum(attr(x, "D2")))

layout(matrix(1:1, 1))
par(mai=c(1.02,0.82,0.82,0.42))    

plot(Lambda, CV / 1e5, log = "x")

sylvia.chrono <- res[[2]]
rts <- attr(sylvia.chrono, "rates")
summary(rts)

par(mar = c(2, 0, 0, 0))
plot(sylvia.chrono, edge.width = 100*rts, label.offset = .15)
axisPhylo()

write.tree(sylvia.chrono, "sylvia.chrono.tre")

```



###
### Chapter 6
###
8. Ancestral character reconstruction. Discrete characters.

```{r}

load("sylvia.RData")
TR <- read.tree("sylvia.txt_phyml_tree.txt")
mltr.sylvia <- TR[[28]]
mltr.sylvia$tip.label <- taxa.sylvia[mltr.sylvia$tip.label]
mltr.sylvia <- root(mltr.sylvia, "Chamaea_fasciata")
mltr.sylvia <- drop.tip(mltr.sylvia, "Chamaea_fasciata")
plot(mltr.sylvia)
DF <- sylvia.eco[mltr.sylvia$tip.label, ]
table(DF$geo.range, DF$mig.behav)

syl.er <- ace(DF$geo.range, mltr.sylvia, type = "d")
syl.er
syl.sym <- ace(DF$geo.range, mltr.sylvia, type="d", model="SYM")
syl.sym
anova(syl.er, syl.sym)

mod <- matrix(0, 3, 3)
mod[2, 1] <- mod[1, 2] <- 1
mod[2, 3] <- mod[3, 2] <- 2
mod
syl.mod <- ace(DF$geo.range, mltr.sylvia, type="d", model=mod)
syl.mod
sapply(list(syl.er, syl.sym, syl.mod), AIC)

Q <- syl.mod$index.matrix
Q
diag(Q) <- 0
Q
Q[1, 2] <- Q[2, 1] <- syl.mod$rates[1]
Q[2, 3] <- Q[3, 2] <- syl.mod$rates[2]
Q

Q <- syl.mod$index.matrix
diag(Q) <- 0
Q
Q[] <- c(0, syl.mod$rates)[Q + 1]
Q
diag(Q) <- -rowSums(Q)
Q


P <- matexpo(0.05 * Q)
#This function computes the exponential of a square matrix using a spectral decomposition. 
rownames(P) <- c("temp", "temptrop", "trop")
colnames(P) <- rownames(P)
P

co <- rep("grey", 24)
co[DF$geo.range == "temp"] <- "black"
co[DF$geo.range == "trop"] <- "white"
plot(mltr.sylvia, type = "c", use.edge.length = FALSE, no.margin = TRUE, label.offset = 1)
tiplabels(pch = 22, bg = co, cex = 2, adj = 1)
nodelabels(thermo = syl.mod$lik.anc, cex = 0.8,
           piecol = c("black", "grey", "white"))


```

9. Diversification analyses
```{r}
sylvia.chrono <- read.tree("sylvia.chrono.tre")
yule(sylvia.chrono)
birthdeath(sylvia.chrono)
1 - pchisq(2*(-7.858507 - -7.858507), 1)

x <- sylvia.eco[sylvia.chrono$tip.label, "geo.range"]
ANC <- ace(x, sylvia.chrono, type = "d", model = mod)
ANC$lik.anc[1:3, ]
anc <- apply(ANC$lik.anc, 1, which.max)
anc
X <- factor(c(x, anc))
yule.cov(sylvia.chrono, ~ X)

1 / (1 + exp(-(0.3200766)))
1 / (1 + exp(-(0.3200766 -2.0038455)))
1 / (1 + exp(-(0.3200766 -1.8433086)))

#bootstrapping
source("yule.cov2.R")
yule.cov2(sylvia.chrono, ~ X)
fsamp <- function(x) sample(length(x), size = 1, prob = x)
nrep <- 1e3
Pvls <- numeric(nrep)
for (i in 1:nrep) {
    anc <- apply(ANC$lik.anc, 1, fsamp)
    X <- factor(c(x, anc))
    Pvls[i] <- yule.cov2(sylvia.chrono, ~ X)
}
hist(Pvls, freq = FALSE, main = "")
lines(density(Pvls))

```











